Turbo fast no compromise gravitational wave parameter estimation

### 2022-11-22 src/tex/ms.tex
Add gradient based MCMC paragraph

Compared to direct sampling, MCMC algorithms only explore regions that are
highly probable, thus reducing the computational cost by not wasting resource in
region where it is unlikely to generate the data.

\texttt{Jax} allows us to compute the gradient of the likelihood function with
respect to the parameters through automatic differentiation. This enables us to
use gradient-based samplers such as Metropolis-adjusted Langevin algorithm
(MALA) \kw{cite} and Hamiltonian Monte Carlo (HMC)\kw{cite}. The main advantage
of gradient-based sampling methods over gradient-free sampling methods is the
gradient information help decorrelates different parameters. This means
gradient-based sampling methods allow longer jump in a high dimensional space.
This is crucial for high dimensional problems \footnote{Gradient-free methods
often have troubles dealing with problems with $>20$ dimensions.}, since
gradient-free methods would need small step size to have reasonable acceptance
rate in a high dimensional space, which in turn leads to a more correlated
samples hence slower convergence.

>>>>

Compared to direct sampling, MCMC algorithms only explore regions that are
highly probable, thus reducing the computational cost by not wasting resource in
region where it is unlikely to generate the data. However, MCMC algorithms come
with their own set of issue. To illustrate what difficulties a MCMC algorithms
may face, we can examine one of the most vanilla MCMC algorithm, the
Metropolis-Hastings algorithm with a Gaussian kernel. Starting at some initial
point, one can draw a proposed point from a Gaussian transition kernel, defined
as
\begin{align}
    q(\mathbf{x},\mathbf{x_0})= \mathcal{N}(\mathbf{x}|\mathbf{x_0},\mathbf{C}),
\end{align}
where $\mathbf{x_0}$ is the current location of the chain,
$\mathbf{x}$ is the proposed location, and $\mathbf{C}$ is the
covariance matrix of the Gaussian. In the simplest case, we can pick
$\mathbf{C}$ to be a diagonal matrix with a constant value, which corresponds to
an isotropic Gaussian centers around the current location with a constant variance.
And the acceptance criteria is defined as
\begin{align}
\alpha(\mathbf{x},\mathbf{x_0}) = \min\left(1,\frac{p(\mathbf{x})q(\mathbf{x_0},\mathbf{x})}{p(\mathbf{x_0})q(\mathbf{x},\mathbf{x_0})}\right).
\label{eq:Gaussian_acceptance}
\end{align}

We can see from eq. \ref{eq:Gaussian_acceptance} the acceptance rate is
proportion to fraction of volume where the posterior density at the proposed
location is higher than the current location within the Gaussian transition
kernel. If we choose the variance of the transition kernel to be too large, this
fraction will be small hence the acceptance rate will be poor. On the other
hand, if one choose the variance to be too small, nearby samples will be
correlated. In both cases, the efficiency in constructing the chain with a
target number of independent samples is suboptimal. There is often a tuning
process before we run the MCMC algorithm to find the optimal tuning parameters,
in this example the variance of the Gaussian, to ensure close to optimal
performance.

However, as we are dealing with higher dimensional problems, even the optimally
tuned Gaussian transition kernel does not guarantee good performance. In order
to have a reasonable acceptance rate, the variance of the Gaussian has to be
smaller in a higher dimensional space, which means the transition kernel is in
general making smaller and smaller step as we increase the dimensionality.

Transition kernels which leverage gradient information of the target
distribution can help to address this issue of shortening steps in a high
dimensional space. Instead of proposing a new point by drawing from a Gaussian,
one can use the gradient evaluate at the current location to propose a new
point. For example, Metropolis-adjusted Langevin algorithm (MALA) \kw{cite}
place a unit Gaussian at the tip of the gradient vector at the current position,
\begin{align}
    \mathbf{x} = \mathbf{x_0} + \tau \nabla\log{p(\mathbf{x_0})} + \sqrt{2\tau}N(0,\mathbf{I}),
\end{align}
where $\tau$ is the step size, which is a tuning parameter. Compared to a
Gaussian centers at the current location, the MALA transition kernel is more
likely to propose a point into the higher posterior density region because of
the gradient term, which helps to boost the acceptance rate.

While transition kernels that use gradient information can help to improve the
acceptance rate, computing the gradient of the posterior density function
introduce additional computational cost, which is not necessary beneficial in
terms of sampling time. If one wants to compute the gradient information through
finite differencing, the additional computation cost goes as at least $\sim
\mathcal{O}(2n)$, where $n$ is the dimension of the problem. \texttt{Jax} allows
us to compute the gradient of the likelihood function with respect to the
parameters through automatic differentiation, which gives the gradient
information down to machine precision at around the same order of time compared
to evaluating the posterior itself. Having access to gradient information
through automatic differentiation is crucial to make the trade-off between using
gradient-based transition kernels and additional computation cost favorable.
